---
title: "Dada2 pipeline for amplicon sequences"
subtitle: "Template version {{< meta version >}}"
version: "1.0"
author:
    name: "Marko Suokas"
    affiliation: "BCO Sequencing Center"
format:
  html:
    theme: sandstone  # or any other quarto theme
    code-tools: true
    code-fold: true
    code-summary: "Show code"
params:
  amplicon: "REQUIRED"
editor: visual
---

```{r dynamic_theming, results='asis', echo=FALSE, message=FALSE, warning=FALSE}

library(bslib)

# 1. Read the theme name from the YAML metadata
theme_setting <- rmarkdown::metadata$format$html$theme
if (is.list(theme_setting) || length(theme_setting) > 1) {
  theme_name <- theme_setting[[1]]
} else {
  theme_name <- theme_setting
}

# 2. Create a bslib theme object to extract variables
if (is.null(theme_name) || theme_name == "default") {
    t <- bs_theme(version = 5)
} else {
    t <- bs_theme(version = 5, bootswatch = theme_name)
}

# 3. Helper to extract colors safely
get_color <- function(var_name, fallback = "#000000") {
  val <- bs_get_variables(t, var_name)
  if (is.na(val)) return(fallback)
  return(unname(val))
}

# 4. Save extracted colors to a globally accessible list
theme_colors <- list(
  primary   = get_color("primary", "#0d6efd"),
  secondary = get_color("secondary", "#6c757d"),
  success   = get_color("success", "#198754"),
  warning   = get_color("warning", "#ffc107"),
  info      = get_color("info", "#0dcaf0"),
  light     = get_color("light", "#f8f9fa"),
  dark      = get_color("dark", "#212529"),
  active    = get_color("component-active-bg", get_color("primary", "#0d6efd")), 
  default   = get_color("default", get_color("body-color", "#212529"))
)

# 5. Inject custom CSS for code blocks and inline code
cat(sprintf("
<style>
  /* Code Blocks: primary-colored left border */
  div.sourceCode { 
    border-left: 4px solid %1$s !important; 
    border-radius: 4px !important; 
  }
  
  /* Inline Code: primary-colored text with light background */
  p code, li code, td code, th code { 
    background-color: %2$s !important; 
    color: %1$s !important; 
    padding: 0.2em 0.4em !important; 
    border-radius: 3px !important; 
  }
</style>
", theme_colors$primary, theme_colors$light))
```

The workflow is adapted from the "Dada2 pipeline guide" and the "Bioconductor Workflow for Microbiome Data Analysis: from raw reads to community analyses" by Benjamin Callahan.

The template requires a metadata file in TSV format (with a minimum of two columns of information) in the root directory and a demultiplexed FASTQ dataset in the reads subdirectory. Confirm the location and the naming of the taxonomy reference files. The simplest way to execute the template is to render the file with Quarto via the command line with the amplicon parameter. For example:

`quarto render dada2_template.qmd -P amplicon:16SV4`

Please note that the example analysis will require you to change variables to correspond to your project metadata. Alternatively, you can disable the execution of code chunks with the eval = FALSE statement.

The first step is setting global parameters and loading libraries.

```{r global_parameters, message = FALSE, warning = FALSE}

# Global options

knitr::opts_chunk$set(warning = FALSE, message = FALSE, echo = TRUE)
knitr::opts_chunk$set(cache = FALSE)

# Amplicon target list

AMPLICON_REGISTRY <- list (
    "16SV3V4" = list (
        fwd = "CCTACGGGNGGCWGCAG",
        rev = "GACTACHVGGGTATCTAATCC",
        type = "16S",
        truncLen = c(270, 230),
        assign_taxonomy = TRUE,
        ref_db = "~/reference/silva_nr99_v138.2_toGenus_trainset.fa.gz",
        assign_species = TRUE,
        species_db = "~/reference/silva_v138.2_assignSpecies.fa.gz",
        strip_prefix = FALSE,
        min_length = 200,
        create_tree = TRUE),
    "16SV4" = list (
        fwd = "GTGYCAGCMGCCGCGGTAA",
        rev = "GGACTACNVGGGTWTCTAAT",
        type = "16S",
        truncLen = c(230,230),
        assign_taxonomy = TRUE,
        ref_db = "~/reference/silva_nr99_v138.2_toGenus_trainset.fa.gz",
        assign_species = TRUE,
        species_db = "~/reference/silva_v138.2_assignSpecies.fa.gz",
        strip_prefix = FALSE,
        min_length = 200,
        create_tree = TRUE),
    "16SV4V5" = list (
        fwd = "GTGYCAGCMGCCGCGGTAA",
        rev = "CCGTCAATTCMTTTGAGTTT",
        type = "16S",
        truncLen = c(260,260),
        assign_taxonomy = TRUE,
        ref_db = "~/reference/silva_nr99_v138.2_toGenus_trainset.fa.gz",
        assign_species = TRUE,
        species_db = "~/reference/silva_v138.2_assignSpecies.fa.gz",
        strip_prefix = FALSE,
        min_length = 200,
        create_tree = TRUE),
    "ITS2" = list (
        fwd = "GTGARTCATCGARTCTTTG",
        rev = "TCCTCCGCTTATTGATATGC",
        type = "ITS",
        truncLen = c(0,0),
        assign_taxonomy = TRUE,
        ref_db = "~/reference/sh_general_release_dynamic_all_19.02.2025.fasta",
        assign_species = FALSE,
        strip_prefix = TRUE,
        min_length = 150,
        allow_overhang = TRUE,
        create_tree = FALSE),
    "ITS1" = list (
        fwd = "CTTGGTCATTTAGAGGAAGTAA",
        rev = "CTGCGTTCTTCATCGAT",
        type = "ITS",
        truncLen = c(0,0),
        assign_taxonomy = TRUE,
        ref_db = "~/reference/sh_general_release_dynamic_all_19.02.2025.fasta",
        assign_species = FALSE,
        strip_prefix = TRUE,
        min_length = 150,
        allow_overhang = TRUE,
        create_tree = FALSE),
    "12STELEO2" = list (
        fwd = "AAACTCGTGCCAGCCACC",
        rev = "GGGTATCTAATCCCAGTTTG",
        type = "12S",
        truncLen = c(130,130),
        assign_taxonomy = FALSE,
        assign_species = FALSE,
        strip_prefix = FALSE,
        min_length = 100,
        create_tree = FALSE),
    "ADCR2" = list (
        fwd = "TGAATGCTCTAGGACCACAC",
        rev = "CGACTAATAAATCCATCTGATAC",
        type = "ADCR2",
        truncLen = c(100,100),
        assign_taxonomy = FALSE,
        assign_species = FALSE,
        strip_prefix = FALSE,
        min_length = 100,
        create_tree = FALSE),
    "RBCL" = list (
        fwd = c("AGGTGAAGTAAAAGGTTCWTACTTAAA", "AGGTGAAGTTAAAGGTTCWTAYTTAAA",
                "AGGTGAAACTAAAGGTTCWTACTTAAA"),
        rev = c("CCTTCTAATTTACCWACWACTG", "CCTTCTAATTTACCWACAACAG"),
        type = "RBCL",
        truncLen = c(250,250),
        assign_taxonomy = FALSE,
        assign_species = FALSE,
        strip_prefix = FALSE,
        min_length = 200,
        create_tree = FALSE)
    )

# Confirm region selection
chosen_amplicon <- params$amplicon

if (is.null(chosen_amplicon) || chosen_amplicon == "REQUIRED") {
  stop("\n[FATAL ERROR]: Execution halted. You must specify a target amplicon parameter.\n",
       "Example: quarto render template.qmd -P amplicon:16SV4\n")
}

if (!chosen_amplicon %in% names(AMPLICON_REGISTRY)) {
  stop(sprintf("\n[FATAL ERROR]: Region '%s' is not defined in the registry.\nAvailable amplicons are: %s\n", 
               chosen_amplicon, paste(names(AMPLICON_REGISTRY), collapse = ", ")))
}

#If it passes, load the configuration
cfg <- AMPLICON_REGISTRY[[chosen_amplicon]]
message("Validation successful. Processing target amplicon: ", chosen_amplicon)

# Assign variables — passed to bash chunks as environment variables
Sys.setenv(FWD_PRIMER = paste(cfg$fwd, collapse = " "))
Sys.setenv(REV_PRIMER = paste(cfg$rev, collapse = " "))
Sys.setenv(MIN_LENGTH = cfg$min_length)

# List of R libraries

library(dada2)
library(mia)
library(vegan)
library(Biostrings)
library(tidyverse)
library(ape)
library(kableExtra)
library(patchwork)
library(ggthemes)
library(ggrepel)
library(ggpubr)
library(ggsci)
library(hrbrthemes)
library(scater)
library(sessioninfo)

# Read metadata file from root directory and set sampleID column as rownames
meta <- "metadata.tsv"

# Find the sample ID column (any letter case) and read only it as character,
# so IDs like "001" keep their leading zeros. Other column types are guessed
meta_cols <- names(read_tsv(meta, n_max = 0, show_col_types = FALSE))
id_col <- meta_cols[tolower(meta_cols) == "sampleid"]
if (length(id_col) != 1) stop("metadata.tsv must contain exactly one 'sampleid' column")

samples_meta <- read_tsv(meta, col_types = setNames(list(col_character()), id_col)) %>%
    rename(sampleid = all_of(id_col)) %>%
    column_to_rownames(var = "sampleid")

#Create directories inside project unless they exist
dir.create("results", showWarnings = FALSE)
dir.create("reads/trimmed", recursive = TRUE, showWarnings = FALSE)

```

```{r parameter_table, echo = FALSE}
# Create dataframe of the chosen amplicon settings
cfg_df <- data.frame(
  Setting = names(cfg),
  Value = sapply(cfg, function(x) paste(x, collapse = ", "))
)
rownames(cfg_df) <- NULL

# Output as kable table
kable(cfg_df, format = "html", caption = paste("Amplicon settings for", chosen_amplicon)) %>%
  kable_styling(bootstrap_options = c("striped", "hover")) %>%
  row_spec(0, background = theme_colors$primary, color = "white")
```

PCR primers must be removed before denoising — any remaining primer sequence will be treated as biological variation and produce spurious ASVs. Cutadapt trims the defined primer pair from both read orientations and discards read pairs where no primer was detected (`--discard-untrimmed`). *Note:* *If **cutadapt** is not installed system-wide, you need to activate the Conda environment in two first lines.*

```{bash trim_primers, eval = TRUE}

source ~/.miniconda3/etc/profile.d/conda.sh
conda activate base

# Define input and output directories
indir="reads"           # directory containing your R1 and R2 fastq.gz files
outdir="reads/trimmed"        # output directory for trimmed reads
mkdir -p "$outdir"

# Start a fresh cutadapt log for this run
logfile="results/cutadapt.log"
: > "$logfile"

# Parse multiplexed primers into bash arrays
FWD_ARGS=()
for p in $FWD_PRIMER; do
    FWD_ARGS+=("-g" "$p")
done

REV_ARGS=()
for p in $REV_PRIMER; do
    REV_ARGS+=("-G" "$p")
done

# Loop through all R1 files and run cutadapt
for r1 in ${indir}/*_R1_*.fastq.gz; do
    # Derive R2 file name
    r2="${r1/_R1_/_R2_}"
    # Extract base name without path
    base=$(basename "$r1" | sed 's/_R1_.*//')

    echo "Processing sample: $base"

    cutadapt \
        "${FWD_ARGS[@]}" \
        "${REV_ARGS[@]}" \
        -o "${outdir}/${base}_R1_trimmed.fastq.gz" \
        -p "${outdir}/${base}_R2_trimmed.fastq.gz" \
        "$r1" "$r2" \
        --discard-untrimmed \
        --minimum-length "$MIN_LENGTH"  \
        --report=minimal >> "$logfile" 2>&1
done

```

### Read processing

#### File and sample lists

Build sorted file path lists for forward and reverse reads and extract sample names. The sort order must match between forward and reverse lists — DADA2 pairs reads by position, not by name. Sample names are extracted from the first underscore-delimited token of the filename, so sample IDs should be simple alphanumeric codes without underscores. Use a separate metadata column for human-readable sample names.

```{r file_lists}

#List sequence files in the source directory
list.files("reads/trimmed")

# Forward read names: SAMPLENAME_PATTERN
fnFwds <- sort(list.files("reads/trimmed",
                          pattern = "_L001_R1_trimmed.fastq.gz",
                          full.names = TRUE))

# Reverse read names: SAMPLENAME_PATTERN
fnRevs <- sort(list.files("reads/trimmed",
                          pattern = "_L001_R2_trimmed.fastq.gz",
                          full.names = TRUE))

# Extract sample names to the list
sample.names <- sapply(strsplit(basename(fnFwds), "_"), `[`, 1)

```

#### Read quality plots

Quality profiles are aggregated across up to 400,000 randomly sampled reads per direction. Inspect these plots to confirm that truncation lengths are suitable for the filter and trim step. The MiSeq i100 retains high quality longer than older sequencing platforms.

```{r read_quality, eval = TRUE}

qfwd <- plotQualityProfile(fnFwds, aggregate = TRUE,
                           n = 4e05)
saveRDS(qfwd, file = "results/qfwd.rds")

qrev <- plotQualityProfile(fnRevs, aggregate = TRUE,
                           n = 4e05)
saveRDS(qrev, file = "results/qrev.rds")

```

Plot forward and reverse read qualities. Data can be used to evaluate proper truncation length for filter and trim step.

```{r q-plot, fig.dim = c(12,7)}

qfwd <- readRDS("results/qfwd.rds")
qrev <- readRDS("results/qrev.rds")

qfwd <- qfwd + ggtitle("Forward reads") + theme_ipsum_rc(base_size = 18)
qrev <- qrev + ggtitle("Reverse reads") + theme_ipsum_rc(base_size = 18)

qfwd + qrev

```

#### Filter and trim reads

Reads are truncated to a fixed length and discarded if they contain ambiguous bases (`maxN=0`), exceed the maximum expected errors threshold (`maxEE=2`), or drop below quality score 2 (`truncQ=2`). Truncation lengths are set in the amplicon settings and should be chosen so that forward and reverse reads still overlap sufficiently for merging — typically at least 20 bp of overlap. PhiX reads are removed if `rm.phix = TRUE`.

Define output file paths for filtered reads.

```{r filter_list}

# Filtered read files are placed in filtered subdirectory
filtFwds <- file.path("reads/trimmed", "filtered",
                      paste0(sample.names, "_F_filt.fastq.gz"))
filtRevs <- file.path("reads/trimmed", "filtered",
                      paste0(sample.names, "_R_filt.fastq.gz"))

```

Run filtering and trimming. Results are saved to disk in the next chunk so this step can be skipped on re-runs by setting `eval = FALSE`.

```{r filterandtrim, eval = TRUE}

# Default parameters except truncLen and multithread
out <- filterAndTrim(fnFwds, filtFwds, fnRevs, filtRevs,
                     truncLen = cfg$truncLen,
                     maxN = 0, maxEE = 2, truncQ = 2,
                     compress = TRUE, multithread = TRUE,
                     rm.phix = TRUE)

```

#### Remove zero-read samples

Check for samples that lost all reads during filtering and remove them from file lists and metadata. The full sample list is kept, so removed samples are still reported in the denoising summary.

```{r zero_read_check, eval = TRUE}

# Read filtering results from file if the filterandtrim chunk was skipped
if (!exists("out")) out <- readRDS("results/out.rds")

# Keep the full sample list for the denoising summary
all_samples <- sample.names
rownames(out) <- all_samples

# Samples with zero or NA reads after filtering will cause errors in downstream steps
zero_samples <- is.na(out[, "reads.out"]) | out[, "reads.out"] == 0

if (any(zero_samples)) {
  cat("Removing", sum(zero_samples), "sample(s) with zero reads after filtering:\n")
  cat(paste(sample.names[zero_samples], collapse = "\n"), "\n")

  filtFwds <- filtFwds[!zero_samples]
  filtRevs <- filtRevs[!zero_samples]
  sample.names <- sample.names[!zero_samples]
  samples_meta <- samples_meta[rownames(samples_meta) %in% sample.names, , drop = FALSE]
} else {
  cat("All samples passed filtering.\n")
}

# Save objects to files
saveRDS(out, "results/out.rds")
saveRDS(all_samples, "results/all_samples.rds")

```

#### Learn error rates

DADA2 learns the instrument-specific error rates from the data by fitting a parametric model to the observed frequency of each possible base transition as a function of quality score. This model is later used to distinguish true biological sequences from sequencing errors during denoising. This pipeline is optimised for the MiSeq i100 platform, which uses binned (discrete) quality scores. The bins `c(2, 9, 23, 38)` are MiSeq i100-specific and must be changed if a different sequencing platform is used or bins are changed in future RTA updates.

```{r binnedvalues}

miseqBinnedErrfun <- makeBinnedQualErrfun(c(2, 9, 23, 38))

```

```{r learn_errors, eval = TRUE}

errFwd <- learnErrors(filtFwds, errorEstimationFunction = miseqBinnedErrfun,
                      multithread = TRUE, randomize = TRUE)

errRev <- learnErrors(filtRevs, errorEstimationFunction = miseqBinnedErrfun,
                      multithread = TRUE, randomize = TRUE)

# save objects to files
saveRDS(errFwd, "results/errFwd.rds")
saveRDS(errRev, "results/errRev.rds")

```

Visualise the learned error rates. The estimated error rates (black line) should follow the observed error frequencies (points) and decrease monotonically with increasing quality score. Poor model fit suggests that the error learning step should be re-run with more reads or that the quality score binning values need adjustment.

```{r plot_errors, fig.dim = c(12,20)}

errFwd <- readRDS("results/errFwd.rds")

# Plot error rate profile for forward reads
plot_fwd <- plotErrors(errFwd, nominalQ=TRUE) + ggtitle ("Forward") +
    theme_ipsum_rc(base_size = 18)

errRev <- readRDS("results/errRev.rds")

# Plot error rate profile for reverse reads
plot_rev <- plotErrors(errRev, nominalQ=TRUE) + ggtitle("Reverse") +
        theme_ipsum_rc(base_size = 18)

plot_fwd/plot_rev

```

#### Denoise reads

The core DADA2 step. The algorithm uses the learned error model to infer the true biological sequences underlying the observed reads, correcting sequencing errors and resolving sequences that differ by as little as one nucleotide. Output is a list of denoised objects, one per sample, for each read direction.

```{r denoise, eval = TRUE}

# Denoise both forward and reverse reads using error rate profiles
dadaFwds <- dada(filtFwds, err = errFwd, multithread = TRUE)
dadaRevs <- dada(filtRevs, err = errRev, multithread = TRUE)

# Save objects to files
saveRDS(dadaFwds, "results/dadaFwds.rds")
saveRDS(dadaRevs, "results/dadaRevs.rds")

```

#### Merge reads

Denoised forward and reverse reads are merged into full-length amplicon sequences. Only read pairs that overlap perfectly (no mismatches in the overlap region by default) are retained. Merging failure — where many reads do not merge — usually indicates insufficient overlap due to truncation lengths that are too aggressive.

```{r merge_paired, eval = TRUE}

dadaFwds <- readRDS("results/dadaFwds.rds")
dadaRevs <- readRDS("results/dadaRevs.rds")

# Merge denoised paired-end reads
merged <- mergePairs(dadaFwds, filtFwds, dadaRevs, filtRevs, verbose = TRUE,
                     trimOverhang = isTRUE(cfg$allow_overhang))

# Save object to a file
saveRDS(merged, "results/merged.rds")

```

#### Build ASV table

Merged sequences from all samples are combined into a sample-by-sequence count matrix. Each unique sequence is an ASV (Amplicon Sequence Variant). The length distribution of ASVs should be inspected to confirm that merging produced sequences of the expected amplicon length.

```{r ASV}

merged <- readRDS("results/merged.rds")

# Create sequence table. Rownames are sequences, colnames samples
seqtab <- makeSequenceTable(merged)

# Dimensions of ASV table
dim(seqtab)

```

Chimeric sequences are artefacts formed when an incomplete PCR extension product primes on a different template, producing a hybrid sequence that is not present in the original sample. `removeBimeraDenovo` detects these by checking whether each ASV can be reconstructed from two more abundant parent sequences. In a typical 16S dataset, chimeras can account for a substantial fraction of unique sequences but a small fraction of total reads.

```{r chimera_removal, eval = TRUE}

# Remove chimeras
seqtab.nochim <- removeBimeraDenovo(seqtab, method = "consensus",
                                    multithread = TRUE, verbose = TRUE)

# Dimensions
dim(seqtab.nochim)

# Save object to file
saveRDS(seqtab.nochim, "results/seqtabnochim.rds")

```

Inspect the distribution of merged ASV lengths. Note that each bar represents the number of unique sequences at that length, not read counts. A tight distribution around the expected amplicon length confirms clean merging. When studying a different amplicon target, adjust `xlim` to match the expected length range.

```{r plot_ASV_length, fig.dim = c(8,6)}

seqtab.nochim <- readRDS("results/seqtabnochim.rds")

# Inspect sequence length distributions
asv_lengths <- data.frame(length = nchar(colnames(seqtab.nochim)))

# Plot histogram
ggplot(asv_lengths, aes(x = length)) +
  geom_histogram(binwidth = 1, fill = theme_colors$primary) +
  labs(title = "ASV length Distribution",
       x = "Merged Length (nt)",
       y = "Frequency") + xlim(50,600) +
  theme_ipsum_rc(base_size = 14)

```

#### Denoising summary

Track read counts through each processing step to identify where reads are lost. Large drops at the filtering step suggest truncation lengths may be too strict. Large drops at merging indicate insufficient overlap. A high chimera fraction can point to primer dimer contamination or poor library quality, but can also occur in samples with low community richness and high sequencing depth — with few unique templates being amplified intensively, the probability of chimera formation during PCR increases.

```{r summary}

# Read all objects from files
out <- readRDS("results/out.rds")
dadaFwds <- readRDS("results/dadaFwds.rds")
dadaRevs <- readRDS("results/dadaRevs.rds")
merged <- readRDS("results/merged.rds")
seqtab.nochim <- readRDS("results/seqtabnochim.rds")

all_samples <- readRDS("results/all_samples.rds")

# Get values for each sample
getN <- function(x) sum(getUniques(x))

# Place values of retained samples into a vector covering all samples.
# Samples removed after filtering get 0
fill_all <- function(values) {
  v <- setNames(numeric(length(all_samples)), all_samples)
  v[sample.names] <- values
  v
}

track <- cbind(out[all_samples, , drop = FALSE],
               fill_all(sapply(dadaFwds, getN)),
               fill_all(sapply(dadaRevs, getN)),
               fill_all(sapply(merged, getN)),
               fill_all(rowSums(seqtab.nochim)),
               fill_all(rowSums(seqtab.nochim != 0)))

# If processing a single sample, remove the sapply calls: e.g.
# replace sapply(dadaFs, getN) with getN(dadaFs)

# Replace column names
colnames(track) <- c("Input", "Filtered", "DenoisedF", "DenoisedR",
                     "Merged", "Non-chimeric","Variants")

# Mark samples removed due to zero reads after filtering
track <- data.frame(track, check.names = FALSE)
track$Removed <- ifelse(rownames(track) %in% sample.names, "", "yes")

# Table
kable(track, format = "html") %>%
  kable_styling(bootstrap_options = c("striped", "hover")) %>%
  row_spec(0, background = theme_colors$primary, color = "white")

```

#### Assign taxonomy

ASVs are classified with a naive Bayesian classifier against the reference database set in the amplicon settings: SILVA for 16S targets and UNITE for ITS targets. Classification is performed with `assignTaxonomy`. For 16S targets, it is followed by exact-match species assignment with `addSpecies`. Species assignment is run in chunks to avoid memory issues with large ASV tables — unassigned ASVs at species level are expected and normal.

```{r taxonomy}
#| eval: !expr cfg$assign_taxonomy

seqtab.nochim <- readRDS("results/seqtabnochim.rds")

# Assign taxa up to genus level
taxa <- assignTaxonomy(seqtab.nochim, cfg$ref_db, multithread = TRUE)

saveRDS(taxa, "results/taxa.rds")

```

Species assignment is executed only for bacterial targets.

```{r species}
#| eval: !expr cfg$assign_species

taxa <- readRDS("results/taxa.rds")

# Keep genus-level ranks only, in case species were already added in an earlier run
taxa <- taxa[, 1:6, drop = FALSE]

chunk_size <- 3000

n_total <- nrow(taxa)

n_chunks <- ceiling(n_total / chunk_size)

cat("Processing", n_total, "taxa in", n_chunks, "chunks of", chunk_size, "\n")

# ---- First chunk (to detect structure) ----
first_end <- min(chunk_size, n_total)
cat("▶️ Pre-scanning first chunk to set structure\n")
taxa_first <- taxa[1:first_end, , drop = FALSE]
taxa_first_species <- addSpecies(taxa_first, cfg$species_db)

# Create empty output matrix with same structure as full addSpecies() output
taxa_species <- matrix(NA_character_, nrow = n_total,
                       ncol = ncol(taxa_first_species))
colnames(taxa_species) <- colnames(taxa_first_species)
rownames(taxa_species) <- rownames(taxa)

# Insert first chunk results
taxa_species[rownames(taxa_first_species), ] <- taxa_first_species

# ---- Remaining chunks ----
if (n_chunks > 1) {
  for (i in 2:n_chunks) {
    start <- (i - 1) * chunk_size + 1
    end <- min(i * chunk_size, n_total)
    cat("▶️ Chunk", i, ":", start, "-", end, "\n")

    taxa_chunk <- taxa[start:end, , drop = FALSE]

    taxa_chunk_species <- tryCatch({
      addSpecies(taxa_chunk, cfg$species_db)
    }, error = function(e) {
      cat("⚠️ Error in chunk", i, ":", conditionMessage(e),
          "\nReturning original chunk\n")
      return(taxa_chunk)
    })

    taxa_species[rownames(taxa_chunk_species),
                 colnames(taxa_chunk_species)] <- taxa_chunk_species
  }
}

cat("✅ Species assignment finished successfully.\n")

saveRDS(taxa_species, "results/taxa.rds")
```

#### TreeSummarizedExperiment object

All data — count matrix, taxonomy table, sample metadata, and representative sequences — are assembled into a `TreeSummarizedExperiment` (TSE) object. This is the primary data structure for downstream analysis with the `mia` ecosystem. Additionally, non-target sequences (non-bacterial/archaeal kingdoms, non-fungal, mitochondria, and chloroplasts) and taxon prefixes are stripped at this stage, if removal is set in the amplicon parameters.

```{r tse, eval = TRUE}

seqtab.nochim <- readRDS("results/seqtabnochim.rds")

# Set representative sequences
repseq <- DNAStringSet(colnames(seqtab.nochim))

# Read and modify taxonomy table
has_taxonomy <- cfg$assign_taxonomy && file.exists("results/taxa.rds")

if (has_taxonomy) {
  taxa <- readRDS("results/taxa.rds")
  rownames(taxa) <- paste0("ASV", seq(nrow(taxa)))
  base_ranks <- c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus")
  colnames(taxa) <- if (ncol(taxa) == 7) c(base_ranks, "Species") else base_ranks
}

# Modify counts table
assay_data <- seqtab.nochim
colnames(assay_data) <- paste0("ASV", seq(ncol(assay_data)))

# Sample names come from the read file names; metadata is matched to them below
rownames(assay_data) <- sample.names
assay_data <- t(assay_data)

# Stop if any sequenced sample is missing from the metadata
meta_idx <- match(colnames(assay_data), rownames(samples_meta))
if (anyNA(meta_idx)) {
  stop("Samples missing from metadata.tsv: ",
       paste(colnames(assay_data)[is.na(meta_idx)], collapse = ", "))
}

# Reorder metadata rows to match assay data column order if necessary
samples_meta <- samples_meta[meta_idx, , drop = FALSE]

# Tse object
tse <- if (has_taxonomy) {
  TreeSummarizedExperiment(assays = list(counts = assay_data),
                            rowData = taxa,
                            colData = samples_meta)
} else {
  TreeSummarizedExperiment(assays = list(counts = assay_data),
                            colData = samples_meta)
}

referenceSeq(tse) <- repseq

# Filter 16S data
if (has_taxonomy && cfg$type == "16S") {
  # Select only bacterial and archaea asvs
  tse <- tse[rowData(tse)$Kingdom %in% c("Bacteria", "Archaea"), ]
  # Filter mitochondria, keep empty at family level
  tse <- tse[!rowData(tse)$Family %in% "Mitochondria" | is.na(rowData(tse)$Family), ]
  # Filter chloroplast, keep empty at order level
  tse <- tse[!rowData(tse)$Order %in% "Chloroplast" | is.na(rowData(tse)$Order), ]
}

# Strip fungal rank prefixes from taxonomy
if (has_taxonomy && cfg$strip_prefix) {
  rowData(tse) <- DataFrame(
    data.frame(rowData(tse)) %>%
      mutate(across(everything(), ~ str_remove(., "^[a-z]__")))
  )
}

# Filter ITS data
if (has_taxonomy && cfg$type == "ITS") {
  tse <- tse[rowData(tse)$Kingdom %in% "Fungi", ]
}

# Save the object
saveRDS(tse, "results/tse.rds")

```

#### Build phylogenetic tree from ASV variants

Tree building is enabled with `create_tree` in the amplicon settings (16S targets). Representative sequences are aligned with MAFFT, a tree is built with FastTree, midpoint rooted and added to the TSE object.

```{r phylo_tree_function}
#| eval: !expr isTRUE(cfg$create_tree)
#
# Uses the widely-accepted MAFFT alignment and FastTree2 combination,
# and applies a midpoint root before adding the tree into the TSE object.
# Designed to scale efficiently to large data sets
#' @param tse A TreeSummarizedExperiment object.
#' @param threads Integer. Number of threads to use for MAFFT alignment (default: 1).
#'        FastTree is run single-threaded, since it is fast enough without
#'        parallelism and threading only takes effect with OpenMP-enabled
#'        FastTree builds (e.g. FastTreeMP), which cannot be assumed present.
#' @param mafft_args Character arguments to MAFFT.
#'        Default "--auto --parttree"
#' @param fasttree_args Character. Arguments for FastTree. 
#'        Default "-nt -gtr -gamma" (nucleotide, GTR+Gamma model).
#' @param seq_accessor Function. Used to extract sequences from the TSE. 
#'        Default is `TreeSummarizedExperiment::referenceSeq`. If you use a custom 
#'        accessor like `repseq`, pass it here (e.g. `seq_accessor = repseq`).
#' @param verbose Logical. If TRUE, prints progress.
#' @return A TreeSummarizedExperiment object with `rowTree` updated.
#' @export

add_phylotree_to_tse <- function(tse, 
                                 threads = 1,
                                 mafft_args = "--auto --parttree",
                                 fasttree_args = "-nt -gtr -gamma",
                                 seq_accessor = TreeSummarizedExperiment::referenceSeq,
                                 verbose = TRUE) {
    
    # 1. Extract sequences
    if (verbose) message("1. Extracting sequences from TSE...")
    seqs <- seq_accessor(tse)
    
    if (is.null(seqs) || length(seqs) == 0) {
        stop("No sequences found in TSE. Check seq_accessor function.")
    }
    
    n_seqs <- length(seqs)
    if (verbose) message(sprintf("   Found %d sequences.", n_seqs))
    
    # 2. Check for dependencies
    
    if (Sys.which("mafft") == "") stop("mafft is not found in PATH.")
    fasttree_bin <- ifelse(Sys.which("FastTree") != "", "FastTree", 
                           ifelse(Sys.which("fasttree") != "", "fasttree", ""))
    if (fasttree_bin == "") stop("FastTree is not found in PATH.")
    
    # 3. Setup tmp files
    
    tmp_dir <- tempdir()
    fasta_in  <- tempfile(pattern = "seqs_", tmpdir = tmp_dir, fileext = ".fasta")
    fasta_out <- tempfile(pattern = "aln_", tmpdir = tmp_dir, fileext = ".fasta")
    tree_out  <- tempfile(pattern = "tree_", tmpdir = tmp_dir, fileext = ".nwk")
    on.exit(unlink(c(fasta_in, fasta_out, tree_out)))
    
    Biostrings::writeXStringSet(seqs, filepath = fasta_in)
    
    # 4. MAFFT alignment
    
    if (verbose) message("2. Running MAFFT alignment...")
    
    mafft_cmd <- sprintf("mafft --thread %d %s %s > %s", 
                         threads, mafft_args, shQuote(fasta_in), shQuote(fasta_out))
    
    if (verbose) message("   Command: ", mafft_cmd)
    res_mafft <- system(mafft_cmd, intern = FALSE, ignore.stdout = FALSE, ignore.stderr = !verbose)
    
    if (res_mafft != 0 || !file.exists(fasta_out) || file.info(fasta_out)$size == 0) {
        stop("MAFFT alignment failed.")
    }
    
    # 5. FastTree
    
    if (verbose) message("3. Running FastTree...")
    
    fasttree_cmd <- sprintf("%s %s %s > %s", 
                            fasttree_bin, fasttree_args, shQuote(fasta_out), shQuote(tree_out))
    
    if (verbose) message("   Command: ", fasttree_cmd)
    res_fasttree <- system(fasttree_cmd, intern = FALSE, ignore.stdout = FALSE, ignore.stderr = !verbose)
    
    if (res_fasttree != 0 || !file.exists(tree_out) || file.info(tree_out)$size == 0) {
        stop("FastTree failed.")
    }
    
    # 6. Load tree, clamp artifacts, then root
    if (verbose) message("4. Reading tree and applying midpoint rooting...")
    tree <- ape::read.tree(tree_out)
    
    if (is.null(tree)) stop("Failed to read tree file.")
    
    # FastTree can produce extremely small negative branch lengths due to
    # heuristics/rounding; clamp BEFORE rooting so root placement isn't
    # distorted by negative path lengths.
    neg_branches <- sum(tree$edge.length < 0, na.rm = TRUE)
    if (neg_branches > 0) {
        max_neg <- min(tree$edge.length, na.rm = TRUE)
        if (verbose) {
            message(sprintf("   Clamping %d negative branch length(s) to 0 (most negative: %.6g).",
                            neg_branches, max_neg))
        }
        if (abs(max_neg) > 0.01) {
            warning(sprintf(
                "Largest negative branch length (%.4g) is unusually large — may indicate a tree/alignment quality issue rather than numerical noise.",
                max_neg))
        }
        tree$edge.length[tree$edge.length < 0] <- 0
    }
    
    tree <- phangorn::midpoint(tree)
    tree <- ape::ladderize(tree)
    
    # 7. Attach tree to TSE
    
    if (verbose) message("5. Attaching tree to TSE...")
    TreeSummarizedExperiment::rowTree(tse) <- tree
    
    if (verbose) message("Pipeline complete!")
    return(tse)
}
```

```{r phylo_tree}
#| eval: !expr isTRUE(cfg$create_tree)

tse <- add_phylotree_to_tse(tse, threads = 4)
```

#### Write final object into files

Samples with zero total counts are removed before export. The TSE object is saved as an RDS file for use in downstream analysis scripts. Flat-file exports (taxonomy, an ASV count table, metadata, and representative sequences as FASTA) are written for compatibility with other tools.

```{r write_files, eval = TRUE}

# Include only samples that have counts

tse <- tse[, colSums(assay(tse)) > 0]

# Save object
saveRDS(tse, "results/tse.rds")

# Save representative sequences as fasta
tse %>% referenceSeq() %>% 
  writeXStringSet("results/repseq.fasta",
                  append = FALSE, compress = FALSE, format = "fasta")

# Save taxonomy table as tsv

if (has_taxonomy) {
taxfile <- as.data.frame(rowData(tse))
taxfile %>% rownames_to_column(var = "Variant") %>% 
  write_tsv(file = "results/taxonomy.tsv")
}

# Save abundance table as tsv
ASV_counts <- as.data.frame(assays(tse)$counts)
ASV_counts %>% rownames_to_column(var= "Variant") %>%
write_tsv(file = "results/asv_table.tsv")

# Save metadata as tsv
metadf <- data.frame(SampleID = rownames(colData(tse)), colData(tse))
write_tsv(metadf, "results/metadata.tsv")

```

#### Example analysis

Beta diversity describes differences in community composition between samples. Bray-Curtis dissimilarity is calculated on relative abundances and ordinated with PCoA (Principal Coordinates Analysis). Samples with similar microbial communities will cluster together in the ordination plot.

```{r beta_diversity, fig.dim = c(8,8), eval = TRUE}

# Metadata column used for grouping in the example plots. Change to match your metadata
group_var <- "Sample_type"

tse <- transformAssay(
    tse, assay.type = "counts", method = "relabundance")

# Run PCoA on relabundance assay with Bray-Curtis distances
tse <- runMDS(tse, FUN = getDissimilarity,
              method = "bray", assay.type = "relabundance",
              name = "MDS_bray")

# Create ggplot object
bplot <- plotReducedDim(tse, "MDS_bray", colour_by = group_var)

# Calculate explained variance
e <- attr(reducedDim(tse, "MDS_bray"), "eig")
rel_eig <- e / sum(e[e > 0])

# Adjust plot object
bplot <- bplot + labs(
    x = paste("PCoA 1 (", round(100 * rel_eig[[1]], 1), "%", ")", sep = ""),
    y = paste("PCoA 2 (", round(100 * rel_eig[[2]], 1), "%", ")", sep = ""),
    colour = group_var) +
  theme_ipsum_rc(base_size=12)
  
bplot$scales$scales <- list()

bplot$layers[[1]]$aes_params$size <- 3

bplot + scale_color_calc()

```

The PCoA plot is coloured by the selected grouping variable. The percentage of variance explained by each axis is shown on the axis labels.

Alpha diversity quantifies species richness and evenness within each sample. Shannon index accounts for both the number of ASVs and their relative abundances — higher values indicate more diverse and evenly distributed communities.

```{r shannon, eval = TRUE}

# Calculate alpha and create df
tse <- addAlpha(tse, assay.type = "counts",
                index = "shannon_diversity")

alpha_table <- data.frame(Sample = rownames(colData(tse)),
                          Shannon = colData(tse)$shannon_diversity)

kable(alpha_table, digits = 2, caption = "Shannon diversity",
      format = "html") %>% 
  kable_styling(bootstrap_options = c("striped", "hover")) %>%
  row_spec(0, background = theme_colors$primary, color = "white")

```

Boxplot of Shannon diversity grouped by the selected grouping variable. Individual sample points are overlaid to show within-group variation.

```{r boxplot, fig.dim = c(7,7), eval = TRUE}

# Plot boxplot
ggplot(as.data.frame(colData(tse)),
       aes(x = .data[[group_var]], y = shannon_diversity,
           fill = .data[[group_var]])) + geom_boxplot() +
  geom_jitter(width = 0.15, size = 2, alpha = 0.7) +
  theme_ipsum_rc(base_size = 14) +
  labs(
    x = group_var,
    y = "Shannon Diversity",
    title = paste("Shannon diversity by", group_var)
  ) + theme(legend.position = "none") + scale_fill_calc()

```

#### Session info

```{r session_info}

sessioninfo::session_info()
```
